---
title: "In Class Activity - Basic Data Types"
author: "YOUR NAME HERE"
format: 
  html:
    code-tools: true
execute:
  echo: true
---

## Overview & Objective

This is the first foray into how to actually use `R` and focuses on understanding some of the most basic data types.

## Student Learning Objectives

At the end of this topic, you should be able to:  

 - Describe characteristics of numeric, character, logical, and date data types.  
 - Use introspection and coercion for basic data types. 
 - Perform operations on included data types.  


## Activity

The following markdown has a set of questions followed by a chunk for you to put in code to answer the question.  Fill them all out as indicated.

:::{.callout-tip}
In the exercises below and hereafter, I may ask you to show the value of a variable (or because we all strive for programming laziness, just say "Show it.").  in these cases use the variables name by itself on a line in the chunk and it will print it out when you render to HTML or DOCX or PDF.

Also, if there is a response to be given for a question, put it in text below the chunk and if you would make it bold or italic to stand out a bit from the rest of the document.
:::

### Numeric Data

What is your favorite number?  Create a variable named `myFavoriteNumber` and assign that value to it.  Show it.



```{r}
myFavoriteNumber <- 42
```


The ancient Egyptians were quite cleaver at engineering and building things.  The [Rhind Papyrus](https://en.wikipedia.org/wiki/Rhind_Mathematical_Papyrus), which dates to 1650 BCE, indicates that they estimated the ratio between the circumference and the diameter of a circle to be 256 divided by 81.  Create a variable named `Rhind_Pi` that estimates the papyrus version of $\pi$.  Show it.

```{r}
Rhind_Pi <- 256/81
Rhind_Pi
```

Compare it to the value of the real version of pi (in `R`, we use `pi` a built-in variable.) Mathematically is the previous version greater or less than what we use now? Show it.

```{r}
diff <- pi - Rhind_Pi 
diff
```

*It is larger in magnitude than our current version of pi*


The function `sqrt()` provides you the square root of a number. Assign a variable named `root2` to hold the value of `sqrt(2)`.  Show the value of this variable and its value by printing it out.

```{r}
root2 <- sqrt(2)
root2
```


What is the value contained in the variable below named `sqrOfNegative2`?  Is this what you would expect for the square of -2? Why or why not?  Answer this in text below the chunk.

```{r}
sqrOfNegative2 <- -2^2
```

**The order of operations here has the exponent operating before the negation operator.  See [this](https://dyerlabteaching.github.io/Basic-Data-Types/slides.html#/order-of-operations-top-to-bottom) slide for a reminder.**


### Character Data

Create variables for your first and last names (and give them proper variable names).

```{r}
first <- "Rodney"
last <- "Dyer"
```


Combine these variables into a single variable `fullName` using `paste()` and then show the value of the variable.

```{r}
fullName <- paste(first,last)
fullName
```


Create a new variable, `properName`, which is your last name, a comma and space, followed by your first name (see the help for `paste()` and in particular the `sep` argument.)  Use the variables above instead of writing out the letters for your names.  Show it.

```{r}
paste( last, first, sep=", " ) -> properName 
properName
```


The function `nchar()` returns the number of characters in a character variable.  Use it on your `properName`.  How many characters are in your name?

```{r}
nchar( properName )
```


### Logical Data 

Create the variables `trueResult` and `falseResult` and assign them the approprite values.

```{r}
trueResult <- TRUE
falseResult <- FALSE
```


Using the variables `trueResult` and `falseResult` to demonstrate logical operators for *equality*, *inequality*, and *negation*.

```{r}
# there are many combinations of how you could show these, here is one.  
#   I'm just looking for your use of ==, !=, and !

trueResult == falseResult
trueResult != falseResult
!trueResult 
```


Using the variables `trueResult` and `falseResult` demonstrate AND operator in R for all pairwise combinations (e.g, for the and operator, show how to evaluate `trueResult`/`trueResult`, `trueResult`/`falseResult`, `falseResult`/`trueResult`, and `falseResult`/`falseResult`)

```{r}
trueResult && trueResult
trueResult && falseResult
falseResult && trueResult
falseResult && falseResult
```

Do the same as the previous question for the appropriate OR operator.

```{r}
trueResult | trueResult
trueResult | falseResult
falseResult | trueResult
falseResult | falseResult
```


In the example above for the square root, the variable `root2` has the square root of the number $2$.  Create a variable called `two` that multiples `root2 * root2`.  

```{r}
two <- root2 * root2
```


Does the new variable if you use the equality operator, `two` equal `2` (test using equality operator do not look at it).

```{r}
two == 2
```

**This is a can of worms, take your best shot at it and we will discuss.**

### Dates

For these next questions, I'm going to use the library `lubridate` like we did in class.  When did you first become a college student?  Create a variable representing this date.

```{r}
library( lubridate )
student <- mdy("01/15/1990", tz="US/Pacific")
student
```


How many days was it between then and now?  Assign this to a variable named `daysAsStudent` and use it as an in-text reference so that it looks like you typed the number of days.  Look up the functions `format` or `prettyNum` to see about fixing the number of decimal places (or in my case to add the thousands separator...🤦‍♂️)


```{r}
now() - student  -> daysAsStudent
```

**I have been a student for `r prettyNum(daysAsStudent, big.mark=",", digits=2)` days.**


